A weighted sampling algorithm to designing RNA sequences with target secondary structure and nucleotide distribution

### 2013-01-08 ISMB/conclusion_ISMB.tex
fixed ISMB headers
conclusion v1

%!TEX root = main_RECOMB.tex
\section{Conclusion}
\label{sec:conclusion}

>>>>

%!TEX root = main_ISMB.tex
\section{Conclusion}

\label{sec:conclusion}

In this article, we described a novel algorithm, \ourprog, for the RNA design problem, i.e. the design of an RNA sequence adopting a predefined secondary structure as minimal free-energy fold.
Implementing a global sampling approach, it optimizes affinity towards the target secondary structure, while granting the user full control over the \GCContent of the resulting sequences.
This extended control does not necessarily induce additional computational demands, and we showed the linear complexity of both the preprocessing stage and the generation of candidate sequences for the design, allowing for the design of larger and more complex secondary structures in a matter of minutes on a single processor. We evaluated the method on a benchmark composed of target secondary structures extracted from the \RNASTRAND database. We observed good overall success rate, with the notable exception of very low targeted \GCContent ($10\%$), and a good entropy within designed candidates.
Finally, we implemented an hybrid approach, using the \RNAinverse software as a post-processing step for unpaired regions. This approach greatly increased the success rate of the method, allowing for the design of highly diverse candidates for almost all of the structures in our benchmark, while largely preserving the targeted \GCContent.

### 2013-01-08 ISMB/methods_ISMB.tex
conclusion v1

\subsection{Postprocessing unpaired regions: A local/global hybrid approach}
\label{subsec:glocal_method}
Due to our simplified our energy model, unpaired regions are not subject to design constraints other than the \GCContent, leading to modest similarities of refolded design candidates with the targeted structure. To improve these performances and test the complementarity  of our global sampling approach with previous contributions based on local search, we used the \RNAinverse software to redesign unpaired regions, specifying a constraint mask to prevent stacking base-pairs from being modified and, if necessary, reetablishing stacking-pair content afterward.
{\em todo: discuss evolution of \GCContent}

>>>>

\subsection{Postprocessing unpaired regions: A local/global hybrid approach}
\label{subsec:glocal_method}
Due to our simplified our energy model, unpaired regions are not subject to design constraints other than the \GCContent, leading to modest probabilities for refolded design candidates to match the targeted structure. To improve these performances and test the complementarity  of our global sampling approach with previous contributions based on local search, we used the \RNAinverse software to redesign unpaired regions. We specified a constraint mask to prevent stacking base-pairs from being modified and, whenever necessary, reetablished the content content a posteriori.

### 2013-01-08 ISMB/results_ISMB.tex
conclusion v1

\subsection{Implementation}
\label{sec:implementation}
Our software, \texttt{Incarnation}, was implemented in {\tt Python~2.7}. We used
\RNAinverse from the \textit{Vienna Package 2.0}~\cite{Hofacker:1994}.
All time benchmarks were run on a one AMD Opteron(tm) Processor 6278  at 2.4 GHz with cache of 512 KB.
 We used as parameter $\beta=15$, the penalty for an invalid base pair.

We present the average times spent by \texttt{Incarnation} and \RNAinverse to generate one sequence
with the required \texttt{C+G} content in Figure~\ref{fig:time}. As expected, the time grows linearly
in function of the length of the structures for \texttt{Incarnation}.  A highly interesting feature
is the huge decrease in time of computation when there is $15\%$ free 
nucleotides inside the loops of the secondary structure.

\begin{figure}[ht!]
	\centering
	\includegraphics[width=\textwidth]{Figures/time_rnastrand_clustered_rnainverse_100samples_fix}
	\caption{Average time in seconds to generate one sequence. We explicitly show 
	the time spent by \texttt{Incarnation} (full line) and \texttt{RNAinverse} (dashed) for various \texttt{C+G} contents. The first plot is in function
	of the length of the structures. The second is in function of the stack
	density (i.e. $2\cdot\#stacks/length$) and the latter in function of 
	the proportion of free nucleotides (i.e. not base paired) inside loops.}
	\label{fig:time}	
\end{figure}

@@

A first criterion is which percentage of sequences have as MFE the target
secondary structures. We are also interested by the number of structures
for which we can design at least one sequence with the desired MFE.
We show the results when using only \texttt{Incarnation} in 
 Figure~\ref{fig:mfe_struct_solved_noinverse}. As mentioned earlier, the high rate of failure
is due to the fact that no selection criterion is applied to
unpaired nucleotides. A local strategy is thus needed.
After processing \texttt{Incarnation} sequences with \RNAinverse to 
locally optimize the free nucleotides, we obtained the results 
shown in Figure~\ref{fig:mfe_struct_solved}. We observed
that length is not a good predictor for the quality of the results. Instead,
a high number of free nucleotides in the structure seems to be a 
good measure of its design hardness. 
On top of that, with a \texttt{C+G} content of $50\%$ and more, at least
one solutions was found for almost all structures. The difficulty of 
designing sequences for targets with a high number of free nucleotides 
 in loops appears clearly in the last subfigure. Even with a high \texttt{C+G} content a our method was not able to generate solutions for a few
 structures.

@@

\begin{figure}[ht!]	
	\centering
	\includegraphics[width=\textwidth]{Figures/mfe_struct_solved}
	\caption{Quality of \texttt{Incarnation+RNAinverse} results. The first row shows the percentage
	of sampled sequences folding into the target. The second shows the 	
	proportion	of structures for which at least one correct sequence was 
	sampled.}
	\label{fig:mfe_struct_solved}	
\end{figure}

To evaluate the global quality of \texttt{Incarnation} sequences, we show
in Figure~\ref{fig:ss_sens} the ratio of well predicted base pairs in the
MFE structure of our sampled sequences. We can observe that, in all cases,
the hardest sequence to design are those with an extremely low \texttt{C+G}
content. As anticipated, those are the sequences with the weakest bonds.
We notice that the most accurate sequences yield a \texttt{C+G} content
of $70\pm 10\%$.

@@

\begin{figure}[ht!]
 	\centering
	\includegraphics[scale=0.45]{Figures/rnastrand_clustered_rnainverse_100samples_struct_sens.png}
	\caption{Structural sensitivity (i.e. $\#$ well predicted base pairs / $\#$ base pairs in target) of the sampled sequences MFE. }
	\label{fig:ss_sens}	
\end{figure}

@@

A desirable feature in sequence design, is to produce samples with a high
diversity. We show in Figure~\ref{fig:nb_sols_entropy} the number of correct
solutions as their average entropy and base pair entropy, since 
\texttt{Incarnation}  only influences the distribution of nucleotides inside 
stacked base pairs. The most constraining factor for generating valid
 solutions seems to be  low \texttt{C+G} content and the percentage of free nucleotides. Our method is able to generate

@@

\subsection{Global sampling, Local search and Glocal approach}

>>>>

\subsection{Implementation}
\label{sec:implementation}
Our software, \ourprog, was implemented in {\tt Python~2.7}. We used
\RNAinverse from the \textit{Vienna Package 2.0}~\cite{Hofacker:1994}.
All time benchmarks were run on a one AMD Opteron(tm) Processor 6278  at 2.4 GHz with cache of 512 KB.
 We used as parameter $\beta=15$, the penalty for an invalid base pair.

We present the average runtimes observed for \texttt{Incarnation} and \RNAinverse to generate one sequence
with the required \texttt{C+G} content in Figure~\ref{fig:time}. As expected, the time grows linearly
in function of the length of the structures for \texttt{Incarnation}.  A highly interesting feature
is the huge decrease in time of computation when there is $15\%$ free 
nucleotides inside the loops of the secondary structure.

\begin{figure}[ht!]
	\centering
	\includegraphics[width=\textwidth]{Figures/time_rnastrand_clustered_rnainverse_100samples_fix}
	\caption{Average time in seconds to generate one sequence. We explicitly show 
	the time spent by \ourprog (full line) and \RNAinverse (dashed) for various \texttt{C+G} contents. The first plot is in function
	of the length of the structures. The second is in function of the stack
	density (i.e. $2\cdot\#stacks/length$) and the latter in function of 
	the proportion of free nucleotides (i.e. not base paired) inside loops.}
	\label{fig:time}	
\end{figure}

@@

A first criterion is which percentage of sequences have as MFE the target
secondary structures. We are also interested by the number of structures
for which we can design at least one sequence with the desired MFE.
We show the results when using only \ourprog in 
 Figure~\ref{fig:mfe_struct_solved_noinverse}. As mentioned earlier, the high rate of failure
is due to the fact that no selection criterion is applied to
unpaired nucleotides. A local strategy is thus needed.
After processing \ourprog sequences with \RNAinverse to 
locally optimize the free nucleotides, we obtained the results 
shown in Figure~\ref{fig:mfe_struct_solved}. We observed
that length is, in general, not a good predictor for the quality of the results. Instead,
a high number of free nucleotides in the structure seems to make the design difficult. 
On top of that, with a \texttt{C+G} content of $50\%$ and more, at least
one solutions was found for almost all structures. The difficulty of 
designing sequences for targets with a high number of free nucleotides 
 in loops appears clearly in the last subfigure. Even with a high \texttt{C+G} content a our method was not able to generate solutions for a few
 structures.

@@

\begin{figure}[ht!]	
	\centering
	\includegraphics[width=\textwidth]{Figures/mfe_struct_solved}
	\caption{Quality of \ourprog+\RNAinverse results. The first row shows the percentage
	of sampled sequences folding into the target. The second shows the 	
	proportion	of structures for which at least one correct sequence was 
	sampled.}
	\label{fig:mfe_struct_solved}	
\end{figure}

To evaluate the global quality of \ourprog sequences, we show
in Figure~\ref{fig:ss_sens} the ratio of well predicted base pairs in the
MFE structure of our sampled sequences. We can observe that, in all cases,
the hardest sequence to design are those with an extremely low \texttt{C+G}
content. As anticipated, those are the sequences with the weakest bonds.
We notice that the most accurate sequences yield a \texttt{C+G} content
of $70\pm 10\%$.

@@

\begin{figure}[ht!]
 	\centering
	\includegraphics[scale=0.45]{Figures/rnastrand_clustered_rnainverse_100samples_struct_sens.png}
	\caption{Structural sensitivity (i.e. $\#$well predicted base/$\#$base pairs in target) of the sampled sequences MFE. }
	\label{fig:ss_sens}	
\end{figure}

@@

A desirable feature in sequence design, is to produce  highly
diverse samples. Figure~\ref{fig:nb_sols_entropy} shows the number of correct
solutions as their average entropy and base pair entropy, since 
\ourprog  only influences the distribution of nucleotides inside 
stacked base pairs. The most constraining factor for generating valid
 solutions seems to be  low \texttt{C+G} content and the percentage of free nucleotides. Our method is able to generate

@@

%\subsection{Global sampling, Local search and Glocal approach}
